---
title: "Gold & Treasury Yields — Stationarity Testing"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## Why Stationarity Matters

Before testing for any long-run relationship between gold and nominal treasury yields, we need 
to establish the order of integration of each series.

A series is said to be **stationary** (I(0)) if it fluctuates around a constant mean 
with constant variance over time. A series is **integrated of order 1** (I(1)) if it 
is non-stationary in level but becomes stationary after taking first differences.

Cointegration — and the VECM that follows — requires both series to be I(1). 
If one series is I(0) and the other I(1), the framework collapses.

The **Augmented Dickey-Fuller (ADF)** test is a standard tool for this. It tests 
the null hypothesis of a unit root (non-stationarity) against the alternative of stationarity.

$$H_0: \text{series has a unit root (non-stationary)}$$
$$H_1: \text{series is stationary}$$

A **p-value below 0.05** leads us to reject $H_0$ and conclude stationarity.

---

## Setup

```{r}
library(arrow)
library(tidyverse)
library(tseries)   # adf.test()

panel <- read_parquet("../data/macro_panel.parquet")
```

---

## ADF Tests — Levels

We first test each series in level. We expect to **fail to reject** $H_0$ — 
i.e. both series should appear non-stationary in level.

```{r}
adf_gold_level <- adf.test(panel$gold)
adf_yield_level <- adf.test(panel$yield)

tibble(
  Series     = c("Gold (level)", "Yield (level)"),
  Statistic  = c(adf_gold_level$statistic, adf_yield_level$statistic),
  P_value    = c(adf_gold_level$p.value, adf_yield_level$p.value),
  Conclusion = c(
    ifelse(adf_gold_level$p.value  < 0.05, "Stationary I(0)", "Non-stationary I(1)"),
    ifelse(adf_yield_level$p.value < 0.05, "Stationary I(0)", "Non-stationary I(1)")
  )
) |>
  knitr::kable(digits = 3)
```

---

## ADF Tests — First Differences

We now test the first differences. We expect to **reject** $H_0$ — 
both differenced series should be stationary.

```{r}
adf_gold_diff  <- adf.test(diff(panel$gold))
adf_yield_diff <- adf.test(diff(panel$yield))

tibble(
  Series     = c("Gold (1st diff)", "Yield (1st diff)"),
  Statistic  = c(adf_gold_diff$statistic, adf_yield_diff$statistic),
  P_value    = c(adf_gold_diff$p.value, adf_yield_diff$p.value),
  Conclusion = c(
    ifelse(adf_gold_diff$p.value  < 0.05, "Stationary I(0)", "Non-stationary I(1)"),
    ifelse(adf_yield_diff$p.value < 0.05, "Stationary I(0)", "Non-stationary I(1)")
  )
) |>
  knitr::kable(digits = 3)
```

---

## Interpretation

Both series are I(1) in level with p-values of 0.990 for Gold and 0.565 for the 10µyears yield. They also are I(0) in first difference with very strong p-values rejections (both at 0,01). 

The condition for cointegration testing is met and we can proceed to the Johansen test in the next step.