---
title: "Gold & Treasury Yields — Cointegration Testing"
format: html
execute:
  echo: true
  warning: false
  message: false
---

## What is Cointegration?

Two I(1) series are said to be **cointegrated** if a linear combination of them is stationary. Intuitively, even though each series wanders randomly on its own, they move together in the long run — they share a common stochastic trend.

In our case, cointegration between gold and the 10Y yield would mean that deviations from the long-run equilibrium relationship are temporary and self-correcting — which is precisely what H1 and H2 require.

------------------------------------------------------------------------

## The Johansen Test

The **Johansen test** is the standard multivariate framework for cointegration. It tests for the number of cointegrating vectors $r$ in a system of I(1) series.

We test sequentially:

$$H_0: r = 0 \quad \text{(no cointegration)}$$ $$H_0: r \leq 1 \quad \text{(at most one cointegrating vector)}$$

Rejecting $r = 0$ and failing to reject $r \leq 1$ confirms exactly one cointegrating relationship — which is what we expect in a bivariate system.

The test comes in two variants:

-   **Trace test** : tests whether the number of cointegrating vectors is less than $r$
-   **Maximum eigenvalue test** : tests whether there are exactly $r$ cointegrating vectors

We report both.

------------------------------------------------------------------------

## Setup

```{r}
library(arrow)
library(tidyverse)
library(urca)

panel <- read_parquet("../data/macro_panel.parquet")

# Bind series into a matrix — required by ca.jo()
y <- cbind(panel$gold, panel$yield)
colnames(y) <- c("gold", "yield")
```

------------------------------------------------------------------------

## Lag Selection

The Johansen test requires specifying the number of lags. We use the **Akaike Information Criterion (AIC)** to select the optimal lag length.

```{r}
library(vars) 

lag_select <- VARselect(y, lag.max = 10, type = "const")
optimal_lag <- lag_select$selection["AIC(n)"]

cat("Optimal lag (AIC):", optimal_lag, "\n")
```

------------------------------------------------------------------------

## Johansen Test — Trace

```{r}
joh_trace <- ca.jo(
  y,
  type  = "trace",
  ecdet = "const",
  K     = optimal_lag
)

summary(joh_trace)
```

------------------------------------------------------------------------

## Johansen Test — Maximum Eigenvalue

```{r}
joh_eigen <- ca.jo(
  y,
  type  = "eigen",
  ecdet = "const",
  K     = optimal_lag
)

summary(joh_eigen)
```

------------------------------------------------------------------------

## Interpretation

For each test, we compare the **test statistic** to the **critical value at 5%**:

-   For the Trace test, for $H_0: r = 0$ , the test-statistic is of 17.39, which is below the critical value at 5% (19.96) and even below the critical value at 10% (17.85).
-   For the Eigenvalue test, also for $H_0: r = 0$ , the test-statistic is of 13.24, which is also below the critical values at 5% (15.67) and at 10% (13.75).

**We** **expected this first H0 to be rejected**, but in both cases we fail to reject it.

## Conclusion

The cointegration test has failed on the full time window. Still, by looking at the descriptive statistics, we see graphical evidences of a potential inflexion point around 2022 in the relationship of the two curves. We will therefore re-investigate the cointegration between 2011 and 2022.
