---
title: "Gold & Treasury Yields — Cointegration Testing (Sub-period 2011–2021)"
format: html
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Motivation

The full-period Johansen test (2011–2026) failed to confirm
cointegration between gold and the 10Y nominal treasury yield. However,
visual inspection of the raw series suggests a structural break around
2022, when gold and yields began moving simultaneously — a pattern
inconsistent with the classical inverse relationship.

This is corroborated by J.P. Morgan Private Bank (2026)[^1], which
documents a post-2022 asymmetric regime where gold resists yield
increases while still responding to yield decreases.

[^1]: <https://privatebank.jpmorgan.com/eur/en/insights/markets-and-investing/is-it-a-golden-era-for-gold>

We therefore re-run the Johansen test on the **2011–2021 sub-period**,
isolating the classical regime before the structural break.

------------------------------------------------------------------------

## Setup

```{r}
library(arrow)
library(tidyverse)
library(urca)
library(vars)

panel <- read_parquet("../data/macro_panel.parquet")

# Filter to pre-break period
panel_pre <- panel |>
  filter(date < as.Date("2022-01-01"))

cat("Sub-period :", format(min(panel_pre$date)), 
    "→", format(max(panel_pre$date)), "\n")
cat("Observations :", nrow(panel_pre), "\n")

y_pre <- cbind(panel_pre$gold, panel_pre$yield)
colnames(y_pre) <- c("gold", "yield")
```

------------------------------------------------------------------------

## Lag Selection

```{r}
lag_select_pre <- VARselect(y_pre, lag.max = 10, type = "const")
optimal_lag_pre <- lag_select_pre$selection["AIC(n)"]

cat("Optimal lag (AIC):", optimal_lag_pre, "\n")
```

------------------------------------------------------------------------

## Johansen Test — Trace

```{r}
joh_trace_pre <- ca.jo(
  y_pre,
  type  = "trace",
  ecdet = "const",
  K     = optimal_lag_pre
)

summary(joh_trace_pre)
```

------------------------------------------------------------------------

## Johansen Test — Maximum Eigenvalue

```{r}
joh_eigen_pre <- ca.jo(
  y_pre,
  type  = "eigen",
  ecdet = "const",
  K     = optimal_lag_pre
)

summary(joh_eigen_pre)
```

------------------------------------------------------------------------

## Conclusion

The Johansen test fails to confirm cointegration even on the pre-2022
sub-period. Both the Trace statistic (12.50) and the Maximum Eigenvalue
statistic (10.94) fall below their respective 5% critical values (19.96
and 15.67), and even below the 10% thresholds.

This result is analytically meaningful, and consistent with the
theoretical limitation identified in the introduction: the nominal yield
embeds both the real rate and inflation expectations (breakeven
inflation rate), two forces with opposing effects on gold. Even before
2022, this conflation is likely sufficient to prevent a stable
cointegrating vector from emerging.

A static long-run equilibrium framework is therefore inappropriate when
working with the nominal yield as a proxy for gold's opportunity cost.
We consequently abandon the VECM specification and pivot to a **rolling
OLS** approach, which is better suited to capturing a relationship whose
strength and direction evolve over time — and whose instability is
itself the central finding of this analysis.
